---
title: "Do burdened communities sit near major emitters?"
subtitle: "Terrain advanced project · R · CalEnviroScreen + Climate TRACE"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Do burdened communities sit near major emitters?

**Terrain advanced project · R · CalEnviroScreen + Climate TRACE**

Join CalEnviroScreen tract scores with Climate TRACE facility locations to test whether high-burden neighborhoods cluster near large point sources.

**You will build:** a map of tracts and emitters, a buffer-distance analysis, and a short interpretation.

---

## Learning objectives

- Combine environmental justice scores with an emissions inventory
- Compute distances between census tract centroids and facilities
- Compare groups with maps and simple models
- Explain why proximity ≠ causation

**Time:** ~3 hr · [CalEnviroScreen](../explainer.html?id=calenviroscreen) · [Climate TRACE](../explainer.html?id=climate-trace)

---

## For mentors

- **Open in:** RStudio or Posit Cloud.
- **Prerequisites:** CalEnviroScreen beginner project; basic ggplot comfort.
- **Data:** Live CES from ArcGIS + real Climate TRACE CO₂ extract (2022).
- **Watch for:** Students treating correlation as proof that facilities *cause* burden scores.

---

## 1. Setup

```{r packages}
required <- c("jsonlite", "httr", "dplyr", "ggplot2", "readr", "broom", "scales", "geosphere")
missing <- required[!vapply(required, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1))]
if (length(missing)) install.packages(missing, repos = "https://cloud.r-project.org")
library(jsonlite)
library(httr)
library(dplyr)
library(ggplot2)
library(readr)
library(broom)
library(scales)
library(geosphere)
```

**Mentor check-in:** *Why do we need tract **centroids** instead of just county averages?*

---

## 2. Load Climate TRACE facilities (California)

**Goal:** Import georeferenced facility-level CO₂ for 2022.

**You should see:** ~330 facilities with `lat`, `lon`, and `emissions_co2e_tonnes`.

```{r facilities}
GITHUB_RAW <- "https://raw.githubusercontent.com/KrishM23/Environmental-Data-Science/main/notebooks/data"

load_facilities <- function() {
  paths <- c("data/climate_trace_ca_facilities.csv", "notebooks/data/climate_trace_ca_facilities.csv")
  for (p in paths) if (file.exists(p)) return(read_csv(p, show_col_types = FALSE))
  read_csv(paste0(GITHUB_RAW, "/climate_trace_ca_facilities.csv"), show_col_types = FALSE)
}

facilities <- load_facilities() |>
  mutate(
    lat = as.numeric(lat),
    lon = as.numeric(lon),
    emissions_co2e_tonnes = as.numeric(emissions_co2e_tonnes)
  ) |>
  filter(!is.na(lat), !is.na(lon), emissions_co2e_tonnes > 0)

TOP_N_EMITTERS <- 40L
BUFFER_KM <- 10

top_emitters <- facilities |>
  slice_max(emissions_co2e_tonnes, n = TOP_N_EMITTERS, with_ties = FALSE)

message("Loaded ", nrow(facilities), " CA facilities · analyzing top ", nrow(top_emitters), " emitters")
head(top_emitters |> select(source_name, sector, emissions_co2e_tonnes, lat, lon))
```

---

## 3. Load CalEnviroScreen (Los Angeles County)

**Goal:** Pull CES 4.0 scores and tract centroids for LA County (`6037` tracts).

**Concept:** We use ArcGIS twice — once for scores, once for polygon centroids.

```{r ces}
CES_URL <- paste0(
  "https://services1.arcgis.com/PCHfdHz4GlDNAhBb/arcgis/rest/services/",
  "CalEnviroScreen_4_0_Results_/FeatureServer/0/query"
)

fetch_ces_attributes <- function() {
  rows <- list()
  offset <- 0L
  repeat {
    q <- list(
      where = "TractTXT LIKE '6037%'",
      outFields = "TractTXT,CIscoreP,PollutionScore,PopCharScore,pmP",
      returnGeometry = "false",
      resultOffset = offset,
      resultRecordCount = 2000,
      f = "json"
    )
    resp <- GET(CES_URL, query = q)
    stop_for_status(resp)
    parsed <- fromJSON(content(resp, as = "text", encoding = "UTF-8"))
    batch <- parsed$features$attributes
    if (is.null(batch) || nrow(batch) == 0) break
    rows[[length(rows) + 1]] <- batch
    if (!isTRUE(parsed$exceededTransferLimit)) break
    offset <- offset + nrow(batch)
  }
  bind_rows(rows)
}

centroid_from_rings <- function(rings) {
  mat <- do.call(rbind, rings)
  c(lon = mean(mat[, 1]), lat = mean(mat[, 2]))
}

fetch_ces_centroids <- function() {
  rows <- list()
  offset <- 0L
  repeat {
    q <- list(
      where = "TractTXT LIKE '6037%'",
      outFields = "TractTXT",
      returnGeometry = "true",
      outSR = 4326,
      resultOffset = offset,
      resultRecordCount = 500,
      f = "json"
    )
    resp <- GET(CES_URL, query = q)
    stop_for_status(resp)
    parsed <- fromJSON(content(resp, as = "text", encoding = "UTF-8"))
    feats <- parsed$features
    if (is.null(feats) || nrow(feats) == 0) break
    for (i in seq_len(nrow(feats))) {
      rings <- feats$geometry[[i]]$rings[[1]]
      cc <- centroid_from_rings(rings)
      rows[[length(rows) + 1]] <- tibble(
        TractTXT = feats$attributes$TractTXT[i],
        lon = cc[["lon"]],
        lat = cc[["lat"]]
      )
    }
    if (!isTRUE(parsed$exceededTransferLimit)) break
    offset <- offset + nrow(feats)
  }
  bind_rows(rows)
}

ces_scores <- fetch_ces_attributes()
ces_centroids <- fetch_ces_centroids()

tracts <- ces_centroids |>
  inner_join(ces_scores, by = "TractTXT") |>
  mutate(
    CIscoreP = as.numeric(CIscoreP),
    PollutionScore = as.numeric(PollutionScore),
    PopCharScore = as.numeric(PopCharScore)
  )

message("LA County tracts with scores + centroids: ", nrow(tracts))
head(tracts |> select(TractTXT, CIscoreP, lon, lat))
```

**Mentor check-in:** *Why use LA County tracts when emitters are statewide?*

---

## 4. Buffer analysis

**Goal:** For each tract, find distance to the **nearest top emitter** and flag tracts within `BUFFER_KM`.

```{r buffers}
emitter_coords <- top_emitters |> select(lon, lat) |> as.matrix()
tract_coords <- tracts |> select(lon, lat) |> as.matrix()

dist_m <- distm(tract_coords, emitter_coords, fun = distHaversine)
tracts <- tracts |>
  mutate(
    dist_nearest_km = apply(dist_m, 1, min) / 1000,
    near_major_emitter = dist_nearest_km <= BUFFER_KM
  )

summary_tbl <- tracts |>
  group_by(near_major_emitter) |>
  summarise(
    tracts = n(),
    mean_CIscoreP = mean(CIscoreP, na.rm = TRUE),
    median_CIscoreP = median(CIscoreP, na.rm = TRUE),
    .groups = "drop"
  )

print(summary_tbl)
```

**You should see:** Higher average `CIscoreP` for tracts within the buffer (often — not always perfectly separated).

**Mentor check-in:** *If the difference is small, does that mean facilities don't matter?*

---

## 5. Model

```{r model}
fit <- lm(CIscoreP ~ near_major_emitter + PollutionScore + PopCharScore, data = tracts)
tidy(fit)
```

---

## 6. Map

**Goal:** Plot tract burden (color) and top emitters (size).

```{r map, fig.width=9, fig.height=6}
ggplot() +
  geom_point(
    data = tracts,
    aes(x = lon, y = lat, color = CIscoreP),
    size = 1.6, alpha = 0.75
  ) +
  geom_point(
    data = top_emitters,
    aes(x = lon, y = lat, size = emissions_co2e_tonnes),
    color = "#7B2D26", alpha = 0.85, shape = 17
  ) +
  scale_color_gradient(low = "#C7E0A8", high = "#7B2D26", labels = label_number(suffix = "%")) +
  scale_size_continuous(labels = label_comma(), range = c(2, 8)) +
  coord_fixed(ratio = 1.2, xlim = range(tracts$lon) + c(-0.5, 0.5), ylim = range(tracts$lat) + c(-0.3, 0.3)) +
  labs(
    title = "CalEnviroScreen burden vs. major emitters (California top facilities, LA tracts)",
    subtitle = paste0("Orange triangles = top ", TOP_N_EMITTERS, " emitters · buffer = ", BUFFER_KM, " km"),
    x = "Longitude", y = "Latitude",
    color = "CES percentile",
    size = "CO₂ (t)"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

---

## 7. Final takeaway

**Mentor prompts:**

1. Are tracts within `r BUFFER_KM` km of major emitters higher-burden on average?
2. Why is **distance to a facility** an imperfect measure of exposure?
3. What would you add next (wind direction, PM2.5 monitors, income controls)?

**Mentor rubric:** Names both datasets · Avoids causal overclaim · Names one limitation

**Next:** [Climate TRACE map (Python)](california_emitters_map.ipynb) · [CalEnviroScreen beginner](calenviroscreen_burden.Rmd)
