---
title: "Who carries the pollution burden?"
subtitle: "Terrain beginner project · R · CalEnviroScreen"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Who carries the pollution burden?

**Terrain beginner project · R · CalEnviroScreen**

Use CalEnviroScreen 4.0 to explore which Los Angeles County communities face the highest combined pollution and vulnerability.

**You will build:** a ranked chart of the most-burdened census tracts and a short written takeaway.

---

## Learning objectives

By the end, you should be able to:

- Explain what CalEnviroScreen measures at the census-tract level
- Load official California environmental justice data from a public API
- Rank and visualize communities by burden score
- Describe why CES is a *screening* tool, not individual risk

**Time:** ~1 hr · **Dataset:** [CalEnviroScreen explainer](../explainer.html?id=calenviroscreen)

---

## For mentors

- **Open in:** RStudio or Posit Cloud (Knit or run chunk-by-chunk with `Ctrl/Cmd + Enter`).
- **Prerequisites:** None. First R project on Terrain is fine.
- **Watch for:** Students treating percentile scores like raw pollution units — reinforce that **CIscoreP** is a 0–100 ranking.
- **Flow:** Read each section → student runs chunk → use *Mentor check-in* before continuing.

---

## 1. Setup

**Goal:** Load the R packages we need.

**Concept:** `httr` talks to the web API; `dplyr` wrangles tables; `ggplot2` draws charts.

**Mentor check-in:** *Have you used R before, or is this your first script?*

```{r packages}
required <- c("jsonlite", "httr", "dplyr", "ggplot2", "scales")
missing <- required[!vapply(required, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1))]
if (length(missing)) {
  install.packages(missing, repos = "https://cloud.r-project.org")
}
library(jsonlite)
library(httr)
library(dplyr)
library(ggplot2)
library(scales)
```

**You should see:** Package load messages only (or a one-time install prompt).

---

## 2. Load CalEnviroScreen (Los Angeles County)

**Goal:** Pull CES 4.0 tract scores into a data frame.

**Concept:** California publishes CalEnviroScreen on ArcGIS. Tract IDs starting with `6037` are in **Los Angeles County**. We paginate because there are 2,000+ tracts.

**Run the chunk below.**

**You should see:** A message like `Fetched 2343 LA County tracts` and a preview table with `TractTXT`, `CIscoreP`, etc.

**Mentor check-in:** *What is a census tract — a person, a neighborhood, or something else?*

**If something breaks:** Set `USE_SAMPLE <- TRUE` to use bundled offline data, or check internet access.

```{r fetch}
fetch_ces_la <- function() {
  base <- paste0(
    "https://services1.arcgis.com/PCHfdHz4GlDNAhBb/arcgis/rest/services/",
    "CalEnviroScreen_4_0_Results_/FeatureServer/0/query"
  )
  rows <- list()
  offset <- 0
  repeat {
    q <- list(
      where = "TractTXT LIKE '6037%'",
      outFields = "TractTXT,CIscoreP,PollutionScore,PopCharScore,pmP,dieselP,trafficP",
      returnGeometry = "false",
      resultOffset = offset,
      resultRecordCount = 2000,
      f = "json"
    )
    resp <- httr::GET(base, query = q)
    httr::stop_for_status(resp)
    parsed <- jsonlite::fromJSON(httr::content(resp, as = "text", encoding = "UTF-8"))
    batch <- parsed$features$attributes
    if (is.null(batch) || nrow(batch) == 0) break
    rows[[length(rows) + 1]] <- batch
    if (!isTRUE(parsed$exceededTransferLimit)) break
    offset <- offset + nrow(batch)
  }
  dplyr::bind_rows(rows)
}

USE_SAMPLE <- FALSE  # Mentor: set TRUE if offline

if (USE_SAMPLE) {
  sample_path <- if (file.exists("data/ces4_la_county.json")) {
    "data/ces4_la_county.json"
  } else {
    "notebooks/data/ces4_la_county.json"
  }
  ces <- fromJSON(sample_path, flatten = TRUE) |> as_tibble()
  message("Using bundled sample (", nrow(ces), " tracts)")
} else {
  ces <- fetch_ces_la()
  message("Fetched ", nrow(ces), " LA County tracts from ArcGIS")
}

head(ces)
```

---

## 3. Rank the highest-burden tracts

**Goal:** Sort tracts by **CIscoreP** (CalEnviroScreen percentile, 0–100).

**Concept:** Higher percentile = greater combined pollution + population vulnerability relative to all California tracts.

**You should see:** Summary stats and a table of the **top 15** tracts with scores often above the 75th percentile.

**Mentor check-in:** *If a tract has CIscoreP = 90, what does that mean in plain English?*

```{r rank}
ranked <- ces |>
  mutate(
    tract_label = paste0("Tract ", TractTXT),
    CIscoreP = as.numeric(CIscoreP)
  ) |>
  arrange(desc(CIscoreP)) |>
  mutate(rank = row_number())

top15 <- ranked |> slice_head(n = 15)

summary(ranked$CIscoreP)
top15 |> select(rank, TractTXT, CIscoreP, PollutionScore, PopCharScore)
```

---

## 4. Plot

**Goal:** Turn the ranking into a chart you can show someone.

**Concept:** Horizontal bars make long tract labels readable. Color intensity reinforces magnitude.

**You should see:** A bar chart titled *Highest CalEnviroScreen burden in Los Angeles County*.

**Mentor check-in:** *For the #1 tract, is `PollutionScore` or `PopCharScore` larger? What story does that tell?*

```{r plot, fig.width=8, fig.height=5}
ggplot(top15, aes(x = CIscoreP, y = reorder(tract_label, CIscoreP), fill = CIscoreP)) +
  geom_col(show.legend = FALSE) +
  scale_fill_gradient(low = "#D98E5A", high = "#7B2D26") +
  scale_x_continuous(labels = label_number(suffix = "%")) +
  labs(
    title = "Highest CalEnviroScreen burden in Los Angeles County",
    subtitle = "Top 15 census tracts by CIscoreP (CalEnviroScreen 4.0)",
    x = "CalEnviroScreen percentile (CIscoreP)",
    y = NULL,
    caption = "Source: OEHHA CalEnviroScreen 4.0 via California ArcGIS Open Data"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

---

## 5. Final takeaway

Have the student write **3–4 sentences**. Use these prompts:

1. What percentile range do the top tracts fall in?
2. For the #1 tract, is pollution burden or population vulnerability higher?
3. Why is CalEnviroScreen a *screening* tool, not an individual risk score?

**Mentor rubric:** Names specific scores · Uses “percentile” correctly · Mentions screening vs. individual risk

**Next:** [AirNow map (Python)](la_air_quality_map.ipynb) · [Terrain datasets](../datasets.html)
