---
title: "Connect pollution burden to air quality"
subtitle: "Terrain intermediate project · R · CalEnviroScreen + EPA AirNow"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Connect pollution burden to air quality

**Terrain intermediate project · R · CalEnviroScreen + EPA AirNow**

Join CalEnviroScreen burden scores at monitoring locations with EPA AirNow readings to ask: **do higher-burden communities show worse measured air quality?**

**You will build:** a scatter plot and simple linear model linking CES scores to PM2.5 / AQI.

**Before you start**

- Open in **RStudio** or Posit Cloud.
- CalEnviroScreen is fetched live from California's ArcGIS service.
- AirNow defaults to bundled monitor readings; set `USE_LIVE_AIRNOW <- TRUE` and add a free [API key](https://docs.airnowapi.org/account/request/) for live data.

## 1. Setup

```{r packages}
required <- c("jsonlite", "httr", "dplyr", "ggplot2", "readr", "broom")
missing <- required[!vapply(required, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1))]
if (length(missing)) install.packages(missing, repos = "https://cloud.r-project.org")
library(jsonlite)
library(httr)
library(dplyr)
library(ggplot2)
library(readr)
library(broom)
library(scales)
```

## 2. Load AirNow monitor readings

```{r airnow}
USE_LIVE_AIRNOW <- FALSE
AIRNOW_API_KEY <- ""

GITHUB_RAW <- "https://raw.githubusercontent.com/KrishM23/Environmental-Data-Science/main/notebooks/data"

load_airnow_sample <- function() {
  paths <- c("data/airnow_la_sample.json", "notebooks/data/airnow_la_sample.json")
  for (p in paths) {
    if (file.exists(p)) return(fromJSON(p, flatten = TRUE))
  }
  fromJSON(paste0(GITHUB_RAW, "/airnow_la_sample.json"), flatten = TRUE)
}

fetch_airnow_live <- function(api_key, lat = 34.05, lon = -118.25, distance = 75) {
  resp <- GET(
    "https://www.airnowapi.org/aq/observation/latLong/current/",
    query = list(
      format = "application/json",
      latitude = lat,
      longitude = lon,
      distance = distance,
      API_KEY = api_key
    )
  )
  stop_for_status(resp)
  fromJSON(content(resp, as = "text", encoding = "UTF-8"), flatten = TRUE)
}

if (USE_LIVE_AIRNOW && nzchar(AIRNOW_API_KEY)) {
  monitors <- fetch_airnow_live(AIRNOW_API_KEY)
  message("Fetched ", nrow(monitors), " live AirNow observations")
} else {
  monitors <- load_airnow_sample()
  message("Using bundled AirNow sample (", nrow(monitors), " observations)")
}

monitors <- monitors |>
  mutate(
    AQI = as.numeric(AQI),
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude)
  ) |>
  filter(!is.na(AQI), !is.na(Latitude), !is.na(Longitude)) |>
  group_by(SiteName) |>
  slice_max(AQI, n = 1, with_ties = FALSE) |>
  ungroup()

head(monitors |> select(SiteName, ReportingArea, ParameterName, AQI, Latitude, Longitude))
```

## 3. Look up CalEnviroScreen scores at each monitor

For each monitor location we query CalEnviroScreen 4.0 and read the **CIscoreP** percentile for that census tract.

```{r ces_lookup}
CES_URL <- paste0(
  "https://services1.arcgis.com/PCHfdHz4GlDNAhBb/arcgis/rest/services/",
  "CalEnviroScreen_4_0_Results_/FeatureServer/0/query"
)

lookup_ces <- function(lon, lat) {
  resp <- GET(CES_URL, query = list(
    geometry = paste(lon, lat, sep = ","),
    geometryType = "esriGeometryPoint",
    inSR = 4326,
    spatialRel = "esriSpatialRelIntersects",
    outFields = "TractTXT,CIscoreP,PollutionScore,PopCharScore,pmP",
    returnGeometry = "false",
    f = "json"
  ))
  stop_for_status(resp)
  parsed <- fromJSON(content(resp, as = "text", encoding = "UTF-8"))
  attrs <- parsed$features$attributes
  if (is.null(attrs) || nrow(attrs) == 0) return(NULL)
  attrs[1, ]
}

ces_at_monitors <- vector("list", nrow(monitors))
for (i in seq_len(nrow(monitors))) {
  ces <- lookup_ces(monitors$Longitude[i], monitors$Latitude[i])
  if (is.null(ces)) next
  ces_at_monitors[[i]] <- bind_cols(
    monitors[i, c("SiteName", "ReportingArea", "ParameterName", "AQI", "Latitude", "Longitude")],
    ces
  )
}
joined <- bind_rows(ces_at_monitors)
  mutate(
    CIscoreP = as.numeric(CIscoreP),
    PollutionScore = as.numeric(PollutionScore),
    pmP = as.numeric(pmP)
  )

message("Matched ", nrow(joined), " monitors to census tracts")
joined |> select(SiteName, ReportingArea, AQI, CIscoreP, PollutionScore, pmP)
```

## 4. Explore the relationship

```{r model}
fit <- lm(AQI ~ CIscoreP, data = joined)
tidy(fit)

cor_test <- cor.test(joined$CIscoreP, joined$AQI)
cor_test
```

```{r scatter, fig.width=8, fig.height=5}
ggplot(joined, aes(x = CIscoreP, y = AQI)) +
  geom_point(aes(color = ParameterName), size = 3, alpha = 0.85) +
  geom_smooth(method = "lm", se = TRUE, color = "#2F6B4F", fill = "#C7E0A8") +
  scale_x_continuous(labels = label_number(suffix = "%")) +
  labs(
    title = "CalEnviroScreen burden vs. measured air quality",
    subtitle = "Each point is an AirNow monitor; x-axis is CES percentile (CIscoreP) for that tract",
    x = "CalEnviroScreen percentile (CIscoreP)",
    y = "Air Quality Index (AQI)",
    color = "Pollutant",
    caption = "Sources: OEHHA CalEnviroScreen 4.0 · EPA AirNow"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

## 5. Write your takeaway

1. Is the correlation positive, negative, or weak? Does that match your expectations?
2. Why is **correlation not causation** here? (Think monitor placement, weather, timing.)
3. What would you need to strengthen this analysis (daily averages, PM2.5 concentrations, more sites)?

**Next:** Compare with the [beginner CalEnviroScreen notebook](calenviroscreen_burden.Rmd) or the [AirNow map project](la_air_quality_map.ipynb).
