---
title: "Connect pollution burden to air quality"
subtitle: "Terrain intermediate project · R · CalEnviroScreen + EPA AirNow"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Connect pollution burden to air quality

**Terrain intermediate project · R · CalEnviroScreen + EPA AirNow**

Join CalEnviroScreen burden scores at monitoring locations with EPA AirNow readings to ask: **do higher-burden communities show worse measured air quality?**

**You will build:** a scatter plot and simple linear model linking CES scores to AQI.

---

## Learning objectives

- Merge two real environmental datasets with different geography
- Use a spatial lookup (point → census tract)
- Run and interpret a simple regression and correlation
- Explain why correlation ≠ causation in air-quality work

**Time:** ~2 hr · [CalEnviroScreen](../explainer.html?id=calenviroscreen) · [AirNow](../explainer.html?id=epa-airnow)

---

## For mentors

- **Prerequisites:** Beginner CalEnviroScreen project helps; comfort with scatter plots.
- **Two datasets:** CES = long-term burden ranking · AirNow = recent monitor snapshot.
- **Watch for:** Students expecting perfect correlation — real air quality varies hour-to-hour; CES is long-term.
- **Live AirNow:** Requires a free API key — set `AIRNOW_API_KEY` in your environment or paste below.

---

## 1. Setup

**Mentor check-in:** *Why can't we download one CSV that already has both CES and AQI?*

```{r packages}
required <- c("jsonlite", "httr", "dplyr", "ggplot2", "readr", "broom", "scales")
missing <- required[!vapply(required, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1))]
if (length(missing)) install.packages(missing, repos = "https://cloud.r-project.org")
library(jsonlite)
library(httr)
library(dplyr)
library(ggplot2)
library(readr)
library(broom)
library(scales)
```

---

## 2. Load AirNow monitor readings

**Goal:** Build a table of LA-area monitors with latitude, longitude, and AQI.

**Concept:** Each row is a **monitoring site**, not a person. We keep one reading per site (highest AQI if duplicates).

**You should see:** ~15 rows with `SiteName`, `AQI`, `Latitude`, `Longitude`.

**Mentor check-in:** *What is AQI — a pollutant concentration or an index?*

```{r airnow}
AIRNOW_API_KEY <- Sys.getenv("AIRNOW_API_KEY", "")

fetch_airnow_live <- function(api_key, lat = 34.05, lon = -118.25, distance = 75) {
  resp <- GET(
    "https://www.airnowapi.org/aq/observation/latLong/current/",
    query = list(
      format = "application/json",
      latitude = lat,
      longitude = lon,
      distance = distance,
      API_KEY = api_key
    )
  )
  stop_for_status(resp)
  fromJSON(content(resp, as = "text", encoding = "UTF-8"), flatten = TRUE)
}

if (!nzchar(AIRNOW_API_KEY)) {
  stop(
    "AirNow requires a free API key. Get one at https://docs.airnowapi.org/account/request/ ",
    "then set AIRNOW_API_KEY in your R environment or paste it above."
  )
}

monitors <- fetch_airnow_live(AIRNOW_API_KEY)
message("Fetched ", nrow(monitors), " live AirNow observations")

monitors <- monitors |>
  mutate(
    AQI = as.numeric(AQI),
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude)
  ) |>
  filter(!is.na(AQI), !is.na(Latitude), !is.na(Longitude)) |>
  group_by(SiteName) |>
  slice_max(AQI, n = 1, with_ties = FALSE) |>
  ungroup()

head(monitors |> select(SiteName, ReportingArea, ParameterName, AQI, Latitude, Longitude))
```

---

## 3. Look up CalEnviroScreen at each monitor

**Goal:** For each monitor's coordinates, find which census tract it sits in and that tract's **CIscoreP**.

**Concept:** This is a **spatial join** simplified — we ask ArcGIS "which tract contains this point?"

**You should see:** `Matched N monitors to census tracts` and a merged table with both AQI and CIscoreP.

**Mentor check-in:** *Why match monitors to tracts instead of comparing whole counties?*

**If something breaks:** Slow internet can delay lookups (~1 sec per monitor). Sample AirNow still works offline.

```{r ces_lookup}
CES_URL <- paste0(
  "https://services1.arcgis.com/PCHfdHz4GlDNAhBb/arcgis/rest/services/",
  "CalEnviroScreen_4_0_Results_/FeatureServer/0/query"
)

lookup_ces <- function(lon, lat) {
  resp <- GET(CES_URL, query = list(
    geometry = paste(lon, lat, sep = ","),
    geometryType = "esriGeometryPoint",
    inSR = 4326,
    spatialRel = "esriSpatialRelIntersects",
    outFields = "TractTXT,CIscoreP,PollutionScore,PopCharScore,pmP",
    returnGeometry = "false",
    f = "json"
  ))
  stop_for_status(resp)
  parsed <- fromJSON(content(resp, as = "text", encoding = "UTF-8"))
  attrs <- parsed$features$attributes
  if (is.null(attrs) || nrow(attrs) == 0) return(NULL)
  attrs[1, ]
}

ces_at_monitors <- vector("list", nrow(monitors))
for (i in seq_len(nrow(monitors))) {
  ces <- lookup_ces(monitors$Longitude[i], monitors$Latitude[i])
  if (is.null(ces)) next
  ces_at_monitors[[i]] <- bind_cols(
    monitors[i, c("SiteName", "ReportingArea", "ParameterName", "AQI", "Latitude", "Longitude")],
    ces
  )
}

joined <- bind_rows(ces_at_monitors) |>
  mutate(
    CIscoreP = as.numeric(CIscoreP),
    PollutionScore = as.numeric(PollutionScore),
    pmP = as.numeric(pmP)
  )

message("Matched ", nrow(joined), " monitors to census tracts")
joined |> select(SiteName, ReportingArea, AQI, CIscoreP, PollutionScore, pmP)
```

---

## 4. Model the relationship

**Goal:** Test whether higher CES burden aligns with higher measured AQI.

**Concept:** `lm(AQI ~ CIscoreP)` fits a straight line. `cor.test()` gives correlation strength. Neither proves causation.

**You should see:** Regression coefficients and a p-value for correlation.

**Mentor check-in:** *If correlation is weak, does that mean CES is "wrong"?* (No — they measure different things and timescales.)

```{r model}
fit <- lm(AQI ~ CIscoreP, data = joined)
tidy(fit)

cor_test <- cor.test(joined$CIscoreP, joined$AQI)
cor_test
```

---

## 5. Plot

**Goal:** Visualize the relationship for discussion.

**You should see:** Scatter with points colored by pollutant type and a shaded regression band.

**Mentor check-in:** *Point to one outlier — what might explain it?* (Weather, ozone day, monitor near freeway, etc.)

```{r scatter, fig.width=8, fig.height=5}
ggplot(joined, aes(x = CIscoreP, y = AQI)) +
  geom_point(aes(color = ParameterName), size = 3, alpha = 0.85) +
  geom_smooth(method = "lm", se = TRUE, color = "#2F6B4F", fill = "#C7E0A8") +
  scale_x_continuous(labels = label_number(suffix = "%")) +
  labs(
    title = "CalEnviroScreen burden vs. measured air quality",
    subtitle = "Each point is an AirNow monitor; x-axis is CES percentile (CIscoreP) for that tract",
    x = "CalEnviroScreen percentile (CIscoreP)",
    y = "Air Quality Index (AQI)",
    color = "Pollutant",
    caption = "Sources: OEHHA CalEnviroScreen 4.0 · EPA AirNow"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

---

## 6. Final takeaway

**Student writes; mentor reviews:**

1. Is correlation positive, negative, or weak? Expected?
2. Why is **correlation not causation** here?
3. What would strengthen the analysis? (daily averages, PM2.5 µg/m³, more sites, longer time window)

**Mentor rubric:** Mentions both datasets · Names a limitation · Avoids causal overclaim ("CES causes bad air")

**Next:** [CalEnviroScreen beginner notebook](calenviroscreen_burden.Rmd) · [AirNow map](la_air_quality_map.ipynb)
