---
title: "Rebuild the UCLA carbon tracker"
subtitle: "Terrain intermediate project · R · UCLA Emissions Reports"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Rebuild the UCLA carbon tracker

**Terrain intermediate project · R · UCLA Emissions Reports**

Recreate Terrain's homepage carbon tracker: verified UCLA greenhouse gas totals (Scopes 1+2+3), a linear trend from the 2019 baseline, and UC's **90% reduction by 2045** commitment line.

**You will build:** a reproducible emissions chart plus a projected year to reach the 2045 target.

**Before you start**

- Open in **RStudio** or Posit Cloud. Same data and logic as [Terrain's live tracker](../index.html#tracker).
- Read the [UCLA emissions explainer](../explainer.html?id=ucla-emissions).

## 1. Setup

```{r packages}
required <- c("dplyr", "ggplot2", "readr", "scales", "broom")
missing <- required[!vapply(required, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1))]
if (length(missing)) install.packages(missing, repos = "https://cloud.r-project.org")
library(dplyr)
library(ggplot2)
library(readr)
library(scales)
library(broom)
```

## 2. Load verified UCLA emissions

Totals are calendar-year Scopes 1+2+3 from UC Annual Sustainability Reports (same series as `index.html`).

```{r load}
GITHUB_RAW <- "https://raw.githubusercontent.com/KrishM23/Environmental-Data-Science/main/notebooks/data"

load_emissions <- function() {
  paths <- c("data/ucla_emissions.csv", "notebooks/data/ucla_emissions.csv")
  for (p in paths) if (file.exists(p)) return(read_csv(p, show_col_types = FALSE))
  url <- paste0(GITHUB_RAW, "/ucla_emissions.csv")
  read_csv(url, show_col_types = FALSE)
}

emissions <- load_emissions()
emissions
```

```{r meta}
BASELINE_YEAR <- 2019L
BASELINE_EMISSIONS <- 335331
TARGET_YEAR <- 2045L
REDUCTION_PCT <- 0.90
target_emissions <- round(BASELINE_EMISSIONS * (1 - REDUCTION_PCT))

commitment_pace <- function(year) {
  if (year <= BASELINE_YEAR) return(BASELINE_EMISSIONS)
  if (year >= TARGET_YEAR) return(target_emissions)
  frac <- (year - BASELINE_YEAR) / (TARGET_YEAR - BASELINE_YEAR)
  round(BASELINE_EMISSIONS - (BASELINE_EMISSIONS - target_emissions) * frac)
}

cat("2019 baseline:", comma(BASELINE_EMISSIONS), "t CO2e\n")
cat("2045 target (90% cut):", comma(target_emissions), "t CO2e\n")
```

## 3. Fit a trend from the 2019 baseline

```{r trend}
recent <- emissions |> filter(year >= BASELINE_YEAR)
fit <- lm(emissions ~ year, data = recent)
fit_tidy <- tidy(fit)
fit_tidy

slope <- coef(fit)[["year"]]
intercept <- coef(fit)[["(Intercept)"]]
years_ahead <- 2030:2050
projected <- tibble(
  year = years_ahead,
  projected = slope * years_ahead + intercept
)

reach_year <- if (slope >= 0) {
  NA_integer_
} else {
  yr <- ceiling((target_emissions - intercept) / slope)
  if (yr > TARGET_YEAR) yr else NA_integer_
}

if (!is.na(reach_year)) {
  message("Linear trend from 2019 reaches the 2045 target around ", reach_year,
          " (simple extrapolation — not a forecast).")
} else {
  message("At the current post-2019 slope, emissions are not declining toward the 2045 target.")
}
```

## 4. Build the chart

```{r chart, fig.width=9, fig.height=5}
chart_years <- seq(min(emissions$year), TARGET_YEAR)
commitment_df <- tibble(
  year = chart_years,
  commitment = vapply(chart_years, commitment_pace, numeric(1))
)

trend_df <- tibble(
  year = chart_years,
  trend = if_else(year >= BASELINE_YEAR, slope * year + intercept, NA_real_)
)

latest <- emissions |> slice_tail(n = 1)
pace_2023 <- commitment_pace(latest$year)
gap <- latest$emissions - pace_2023

ggplot() +
  geom_line(data = commitment_df, aes(x = year, y = commitment),
            color = "#D98E5A", linewidth = 1, linetype = "dashed") +
  geom_line(data = trend_df, aes(x = year, y = trend),
            color = "#5B8BB0", linewidth = 1) +
  geom_point(data = emissions, aes(x = year, y = emissions),
             color = "#2F6B4F", size = 2.2) +
  geom_vline(xintercept = TARGET_YEAR, color = "#4A5A54", linetype = "dotted", alpha = 0.6) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "UCLA greenhouse gas emissions vs. UC commitment",
    subtitle = paste0("Scopes 1+2+3 · latest ", latest$year, ": ",
                      comma(latest$emissions), " t CO₂e · ",
                      if (gap > 0) paste0(comma(gap), " t above linear pace to 2045") else paste0(comma(abs(gap)), " t below pace")),
    x = "Calendar year",
    y = "Metric tons CO₂e",
    caption = "Sources: UC Annual Sustainability Reports · UC Sustainable Practices Policy (90% by 2045)"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

## 5. Write your takeaway

1. How far is UCLA's latest total from the **2019 baseline** and from the **2045 target**?
2. Is UCLA above or below the **linear commitment pace** in the latest year?
3. Why should you treat the dashed trend line as illustrative, not a prediction?

**Next:** Browse [Terrain's live tracker](../index.html#tracker) or explore [campus programs](../programs.html).
